
---
title: "PHI 420/555 - Legal Probabilism"
subtitle: "Fall 24"
author: "[Marcello Di Bello](https://www.marcellodibello.com/)  - mdibello@asu.edu"
output:
    rmdformats::html_clean:
    fig_width: 6
    fig_height: 6
    highlight: kate
    thumbnails: true
    lightbox: true
    gallery: true
    toc_depth: 2
    toc: TRUE
---    






    






<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"algo-fair-logo.png\" style=\"float: right;width: 300px;height: 220px;\"/>')
   });
</script>



```{r knitr_init, echo=FALSE, results="asis", cache=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print = "75")
opts_chunk$set(echo = FALSE,
	             cache = FALSE,
               prompt = FALSE,
               tidy = TRUE,
               comment = NA,
               message = FALSE,
               warning = FALSE)
opts_knit$set(width = 75)
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<style type="text/css">

body{ /* Normal  */
      font-size: 20px;
      font-family:'Avenir Next';
  }

</style>



<!--- 
ADDED STYLES/ FONT SIZSES
<style type="text/css">
body{ /* Normal  */
      font-size: 18px;
  }
  
 
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>
-->


Th	3:00-5:50 PM in Tempe, [COOR](https://www.asu.edu/map/interactive/?psCode=COOR) 3301  

<!-- ~~3301~~ 3323   --->


Office hours: Thursday after class or by appointment

# Topic

Legal probabilism is an ongoing research program that aims to harness the powers of probability theory to analyze, model and improve the evaluation of evidence and the process of decision-making in trial proceedings.  Despite its promise, significant hurdles exist in carrying out this program, and many have criticized it. These critiques range from difficulties in assessing the probability of someone’s criminal or civil liability to the dehumanization of trial decisions to misconstruing how the process of evidence evaluation and decision-making takes place in trial proceedings. The seminar examines these current debates and their relevance to larger questions about AI and the human future. 

# Objectives

This course is meant for advanced undergraduates and graduate students. 
You will become familiar with current debates in philosophy, law and computer science 
about legal reasoning, evidence and uncertainty with a focus on 
questions about the use of probability, statistics and AI. You will sharpen your critical thinking 
skills, in reading  and  writing, for the analysis of theoretical concepts and their application to current legal questions. You will read academic papers in philosophy, law and computer science and develop an appreciation for their distinctive contributions. 

# Course Materials

Readings and other course materials are available 
on this website. For readings covered 
by copyright, please check the Canvas page of this course.  
Course materials are divided into "primary" and "additional". 
You are only expected to study the primary ones, but I 
recommend that you have a look at the 
additional materials for at least one 
or two weeks.

There is no textbook, but these 
two references can be useful as a guide:

- Urbaniak and Di Bello (2021), [Legal Probabilism](https://plato.stanford.edu/entries/legal-probabilism/)

- [Fundamentals of Probability and Statistical Evidence in Criminal Proceedings](https://www.maths.ed.ac.uk/~cgga/rss.pdf)


# Requirements


## Participation

Please attend class regularly and participate 
(**10% of your grade**). This is a "seminar style" course. You are 
expected to take an active role in the discussions. Please study the assigned 
materials **before class** and be ready to discuss them. 


## Assignments

You should write ten 
pass/fail précis, one pass/fail take-home exam, and three graded essays or 
a research paper (**90% of your grade**).

### (a) Précis

Every week please write a **one-page précis** of one of the papers assigned for that week. 
Unless otherwise noted, the précis should describe: (a) topic of the paper; 
(b) main thesis (or main theses, if there are more than one); 
(c) supporting arguments; (d) objections to these arguments, complications or 
difficulties that the author considers (if any). Submit 
your précis each week through Canvas **before the beginning of class**. 
You should receive a PASS in **at least ten précis**, or else a full letter grade 
will be subtracted from your final grade: A would become B; B would become C; etc.


### (b) Exam

You will take an exam that tests your understanding 
of the basics of probability theory. The exam 
is **take-home** and **pass/fail**.

### (c) Threee essays - option 1 

There will be three main graded assignments 
for this course, **5 pages each**. 

**I**. Write a **critical essay** that outlines the key tenets of legal probabilism and the criticisms 
leveled against it. Your essay should begin with a summary of the key tenets of legal probabilism (1-2 pages). After that, you should apply legal probabilism to a toy example, say a stylized court case (1-2 pages). Finally, you should raise problems and difficulties for legal probabilism (1-2- pages). Please draw from the readings and demonstrate thoughtful engagement with them. 

**II**. After doing independent research, identify a disputed factual issue 
in a court case and analyze it using probability theory. Your essay should serve as **case study** that illustrates the advantages and challenges of applying probability theory to a real legal case. The first part of your essay (1-2 pages) should describe informally the key items of evidence in the case and the disputed facts. The second part (3-4 pages) should apply probability theory to analyze and weigh the evidence. 

**III**. Write a **philosophical argumentative essay** that examines the topic "Legal Probabilism: Opportunities and Challenges". In examining this topic, you should draw from the course readings though I do not expect you to cover all of them. You may (but need not) build on the case study you discussed in your second essay.  Your essay can be primarily negative, e.g., you argue that the challenges are insurmountable. Your essay can also be more positive, e.g., you describe the challenges and propose ways to address them within legal probabilism. 


### (c) Research paper - option 2 

If you are a grad student or advanced undergrad with research experience, 
you may combine the three 5 page essays into one 
**15-20 page** research paper. The research paper should engage closely 
with a subset of the course materials including the additional ones. 
It is neither necessary nor recommended that you use materials outside those already 
included in the course materials. 

*Please come talk to me before you start working on the research paper.* 


# Handouts 

# Schedule 

## 8/22

**Week 1: The Collins case**

Probability and statistical methods can be used in trial proceedings for evaluating 
the evidence presented against the defendant. Sometimes the evidence itself is 
presented in the form of probabilities and statistics. This numerical presentation is often useful but can also 
lead to confusions, exaggerations and errors. To illustrate some of the problems, we will start by 
discussing the infamous case People v. Collins (1968). This case started the law and probability 
debate in the United States back in the 70ies. Along the way, we will learn the basics of 
probability theory.

*Primary readings*

- [People v. Collins](https://www.marcellodibello.com/mathtrial/resources/Collins1968.pdf) (1968)

- Finkelstein and Levin (2001), Statistics for Lawyers, [Sec. 3.1-3.2](https://www.marcellodibello.com/phil90L/resources/FinkelsteinLevin2001Sec3.pdf)

*Additional readings*

- Finkelstein and Fairley (1970), [A Bayesian Approach to Identification Evidence](https://www.marcellodibello.com/mathtrial/resources/FinkelsteinFairley1970.pdf) 

## 8/29

**Week 2: Match evidence**

Match evidence, such as DNA and fingerprint evidence, offers the clearest 
example of how probability and statistics can help to weigh evidence.  We will discuss how 
match evidence works, and the debates about how it should be presented at trial. 
We will continue learning about probability theory.

*Primary readings*

- [People v. Rush](https://www.marcellodibello.com/mathtrial/resources/Rush1995.pdf) (1995)

- Wasserman (2004), [Forensic DNA Typing](https://www.marcellodibello.com/mathtrial/resources/WassermanDNA2008.pdf)

*Additional readings*

- Koehler (1993), [Error and Exaggeration in the Presentation of DNA Evidence](https://www.marcellodibello.com/mathtrial/resources/KoehlerDNAPres1993.pdf)

- Zabell (2005), [Fingerprint Evidence](https://selfteachingresources.pbworks.com/f/Fingerprint+Evidence+jlp13i_zabell.pdf)


##  9/5


**Week 3: Profiling**

There is little doubt that match evidence should be used at trial as evidence of guilt, for example, when the defendant matches the crime traces. But what about evidence that people with certain traits are extremely likely to a commit a crime? For example, data show that people who committed a sexual offense are very likely to commit it again. How should this information be used at trial?


*Primary readings*


<!-- [United States v. Vue](https://casetext.com/case/us-v-vue) (1994)-->

- Pundik (2017), [Freedom and Generalization](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2556690)

- Redmayne (2003), [The Relevance of Bad Character](https://www.cambridge.org/core/journals/cambridge-law-journal/article/abs/relevance-of-bad-character/8876D6D820853C4B38E5914494682BA1)

*Additional readings*

- Aviv (2013), [The Science of Sex Abuse](https://www.marcellodibello.com/phil90L/resources/AvivSexAbuseNewYorker2013.pdf)

- Di Bello and O'Oneil (2020), [Profile Evidence, Fairness, and the Risks of Mistaken Convictions](https://www.journals.uchicago.edu/doi/full/10.1086/705764)


## 9/12


**Week 4: Unlikely coincidences**

If a suspiciously high number of people 
die in a hospital when a nurse is on duty---much higher than when 
other nurses are on duty---is this evidence that 
the nurse is culpable, or was this just a coincidence? 
We will look at two infamous cases, 
Lucia de Berk and Sally Clark, in which statistics about unlikely coincidences
were used---actually, abused!---to conclude that a wrongdoing occurred.

*Primary readings*

- [R v Clark](https://www.marcellodibello.com/mathtrial/resources/SallyClarkSecondAppeal2003.pdf) (2003)

- Derksen (2007), [The Fabrication of Fact](https://www.marcellodibello.com/mathtrial/resources/fabricationoffacts.pdf)

*Additional readings*

- Meester and others (2006), [On the (Ab)use of Statistics in the Legal Case Against the Nurse Lucia de B](https://www.marcellodibello.com/mathtrial/resources/vanLambalgenNurse2006.pdf)

- Fienberg and Kaye (1991), [Legal and Statistical Analysis of some Mysterious Clusters](https://www.marcellodibello.com/mathtrial/resources/FienbergKayeCluster1991.pdf)



## 9/19

**Take-home exam due**

**Week 5: Probability of guilt?**


We have seen how different types of evidence 
can be evaluated using probability and statistics. But what is the 
goal of this evaluation? Some think that the goal is to arrive at 
an assessment of the probability that 
the defendant is guilty, but there are also different probabilistic 
measures of the value of the evidence.


*Primary readings*

- [Regina v. T](https://www.bailii.org/ew/cases/EWCA/Crim/2010/2439.pdf) (2010)

- Dawid (2001), [Bayes’s Theorem and Weighing Evidence by Juries](https://www.researchgate.net/publication/245581914_Bayes's_Theorem_And_Weighing_Evidence_by_Juries)


*Additional readings*

- Fenton (2011), [Improve Statistics in Court](https://www.nature.com/articles/479036a)

- Urbaniak and Di Bello (2021), Legal Probabilism, [Sec. 1 and 2](https://plato.stanford.edu/entries/legal-probabilism/#ProbTool)

## 9/26


**Week 6: Probability and decisions**

Once the evidence has been presented and evaluated, a decision must be made. 
To this end, legal standard of proof serves as decision criteria, for example, convict only if the defendant's guilt was established beyond a reasonable doubt. Since these standards are not precise, legal probabilists have translated them in the more precise language of probability and statistics. The usefulness of this translation is debated because eliminating all reasonable doubts about someone's guilt seems to require more than establishing a high probability of guilt.


*Primary readings*

- Smith (2017), [When does Evidence Suffice for Conviction?](https://academic.oup.com/mind/article/127/508/1193/4555414)

- Gardiner (2019), [The Reasonable and the Relevant: Legal Standards of Proof](https://philarchive.org/rec/GARTRA-9)

*Additional readings*

- Kaplan (1968), [Decision Theory and the Factfinding Process](https://www.jstor.org/stable/1227491)

- Redmayne (2008), [Exploring the Proof Paradoxes](https://www.cambridge.org/core/journals/legal-theory/article/abs/exploring-the-proof-paradoxes/AB54733DCA886ED0BB43156D98F5BBEE)


<!-- Hamer (2004), [Probabilistic Standards of Proof](https://www.marcellodibello.com/innocent/resources/hamer.pdf)-->

<!-- Condorcet -->



## 10/3

**Week 7: Knowledge and legal proof** *(remote class)*

Epistemologists are concerned with knowledge and its value. Knowledge requires more than high probability, as lottery propositions suggest. Could the notion of knowledge help us better understand the standards of proof in legal cases?

*Primary readings*

- Moss (2023), [Knowledge and Legal Proof](https://websites.umich.edu/~ssmoss/Moss%20-%20Knowledge%20and%20Legal%20Proof.pdf)

*Additional readings*

- Papineau (2021), [The Disvalue of Knowledge](https://link.springer.com/article/10.1007/s11229-019-02405-4)




## 10/10

**First essay due**


**Week 8: Relative plausibility**

A prominent alternative to the probabilistic approach is 
the theory of relative plausibility. On this theory, trial decisions 
are guided by comparing the relative plausibility of two competing explanations 
of the evidence. The explanations are put forward by the parties at trial and the 
more plausible should be prevail. This comparative approach tames the complexity 
of trial decision-making in which alternative possibilities about what could 
have happened are potentially infinite in number and cannot be examined in their entirety. 

**Q&A with Ronald Allen**

The first part of the class will consist in a Q&A session with [Ronald Allen](https://www.law.northwestern.edu/faculty/profiles/ronaldallen/), a legal scholar of evidence law, criminal procedure, and constitutional law.  He was recently retained by the Tanzanian Government to assist in the reform of their evidence law. Ronald is the author of one of the readings assigned for today.

*Primary readings*

- Pardo and Allen (2007), [Juridical Proof and Inference and the Best Explanatiopn](https://sites.pitt.edu/~jdnorton/teaching/2682_confirmation/2021/slides/evidence_in_science/Pardo_Allen_2007.pdf)

- Allen (2017), [The Nature of Juridical Proof: Probability as a Tool in Plausible Reasoning](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2982121)

*Additional readings*

- Steele and Stefansson (2021), [Belief Revision for Growing Awareness](https://philarchive.org/rec/STEBRF-6)

- Dahlman (2024), [What Does it Mean to be Plausible?](https://lucris.lub.lu.se/ws/portalfiles/portal/190369771/Dahlman_What_does_it_mean_to_be_plausible_QF_2024_elektroniskt_s_rtryck_.pdf)



## 10/17

**Week 9: The story model**

A close cousin of the theory of relative plausibility is the story model. According to this model, judges and jurors evaluate the evidence presented at trial not in a piecemeal manner, but holistically by constructing a unifying narrative or story of what happened.  Can probability theory model this holistic process? Bayesian networks have been used to calculate numerically the weight of complex bodies of evidence.  We will discuss the prospects of using Bayesian networks for evaluating evidence in trial proceedings.  


**Q&A with Mackor, Dahlman and Lagnado**

The first part of the class will consist in a Q&A session with [Anne Ruth Mackor](https://research.rug.nl/en/persons/anne-ruth-mackor), 
[Christian Dahlman](https://portal.research.lu.se/en/persons/christian-dahlman), and [David Lagnado](https://causalcognitionlab.com/dave/). 
The first two are legal scholars and philosophers, and the third is a cognitive psychologist. They are interested in understanding how judges and jurors assess evidence in conditions of uncertainty. They are the authors of some of the readings assigned for today.


*Primary readings*

- Pennington and Hastie (1993), 'The Story Model for Juror Decision Making', in Inside the Juror: The Psychology of Juror Decision Making - *check canvas*

- Fenton and other (2019), [Analyzing the Simonshaven Case Using Bayesian Networks](https://onlinelibrary.wiley.com/doi/full/10.1111/tops.12417)

*Additional readings*

- Dahlman and Mackor (2019), [Coherence and Probability in Legal Evidence](https://lucris.lub.lu.se/ws/portalfiles/portal/75903137/Coherence_and_probability_in_legal_evidence.pdf)

- Urbaniak and Di Bello (2021), Legal Probabilism, [Sec. 3](https://plato.stanford.edu/entries/legal-probabilism/#BayeNetwForLegaAppl)




## 10/24


**Week 10: The Conjunction Paradox**

Legal cases often require that multiple propositions be established. For example, to establish [bribery](https://www.justice.gov/archives/jm/criminal-resource-manual-2043-comparison-elements-crimes-bribery-and-gratuities), a prosecutor needs to establish several elements: (i) a public official received the bribe; (ii) the bribe was something of value; (iii) the public official carried out an official act because of the bribe; (iv) the bribing party had intent to bribe. Can probability and statistics help to put them together, and yield a unified picture? This is where legal probabilism faces significant challenges, one of the most discussed being the difficulty about conjunction.


**Q&A with Emily Spottswood**

The first part of the class will consist in a Q&A session with [Emily Spottswood](https://law.fsu.edu/faculty-staff/emily-spottswood), a legal scholar interested in how reforms to the rules of procedure and evidence can make civil and criminal litigation more accurate and cost-effective. Emily is the author of one of the readings assigned for today.


*Primary readings*

- Cohen (1977), 'The Difficulty about Conjunction', in The Probable and The Provable - *check Canvas*

- Spottswood (2016), [Unraveling the Conjunction Paradox](https://academic.oup.com/lpr/article/15/4/259/2733187)

*Additional readings*

- Dawid (1987), The Difficulty About Conjunction 

- Haack (2014), Legal Probabilism: An Epistemological Dissent  


## 10/31

**Second essay due**


**Week 11: Beyond the witness?**


When evidence is presented at trial, it will be scrutinized 
and tested against objections by both parties. This cross-examination is 
directed at witnesses, so their live testimonies play 
a central role. But new developments in technology have challenged 
this testimony-centered approach. How to handle machine evidence, such as 
risk assessments, AI diagnostic tools, or softwares that interpret DNA evidence?
Should the testimony-centered approach be abandoned?


**Q&A with Ed Cheng and Alex Nun**

The first part of the class will consist in a Q&A session with [Ed Cheng](https://law.vanderbilt.edu/bio/?pid=edward-cheng), a legal scholar at interested in expert evidence, law and statistics, and remedies, and [Alex Nun](https://law.tamu.edu/faculty-staff/find-people/faculty-profiles/g-alexander-nunn), a legal scholars interested in 
how evidentiary rules and the allocation of decisionmaking authority in the courtroom affect verdict accuracy, efficiency, and legitimacy. Ed and Alex are the authors of one of the readings assigned for today.

*Primary readings*

- Cheng and Nunn (2019), [Beyond the Witness](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3242588)

- Roth (2017), [Machine Testimony](https://www.yalelawjournal.org/pdf/RothFinal_c4o97on1.pdf)

*Additional readings*

- Helmus and others (2022), [Static-99R: Strengths, Limitations, Predictive Accuracy Meta-Analysis, and Legal Admissibility Review](https://psycnet.apa.org/fulltext/2022-61121-001.html)

- Thompson (2023), [Uncertainty in probabilistic genotyping of low template DNA: A case study comparing STRMix and TrueAllele](https://onlinelibrary.wiley.com/doi/full/10.1111/1556-4029.15225)




## 11/07


**Week 12: Eyewitness identifications**

Eyewitness testimony is often the central piece of a 
prosecutor's case against the defendant. Eyewitness evidence is not numerical 
on its face---it consists of qualitative statements "I saw this or that". 
However, a growing body of research in psychology and law has deployed the tools of probability 
and statistics to assess the value of eyewitness evidence. While some studies emphasize its many pitfalls, more recent studies offer a more nuanced picture. 




*Primary readings*

- Loftus and Palmer (1974), [Reconstruction of Auto-mobile Destruction: An example of the Interaction Between Language and Memory](https://www.demenzemedicinagenerale.net/images/mens-sana/AutomobileDestruction.pdf)

- Wixted and Wells (2017), [The Relationship Between Eyewitness Confidence and Identification Accuracy: A New Synthesis](https://journals.sagepub.com/stoken/default+domain/2X3HtwVgmHH7cCakgVgq/full)


*Additional readings*

- Smith et al (2020), [Evaluating the Claim that High Confidence Implies High Accuracy in Eyewitness Identification](https://www.researchgate.net/publication/343485679_Evaluating_the_Claim_that_High_Confidence_Implies_High_Accuracy_in_Eyewitness_Identification)

- Trenary (2013), [State v. Henderson: A Model for Admitting Eyewitness Identification Testimony](https://scholar.law.colorado.edu/lawreview/vol84/iss4/10/)


## 11/14

**Week 13: Wrongful convictions**

We have examined how probability and statistics help to evaluate evidence, piecemeal or in the aggregate, and how they can guide trial decisions. But they can also be used as tools for assessing how 
the trial system performs as a whole, for example, how accurate it is and how often 
innocent people are convicted. Some estimate that the rate of 
wrongful convictions is about 3-5%. 

*Primary readings*

- Risinger (2007), [Innocents Convicted](https://www.marcellodibello.com/innocent/resources/risinger.pdf)


*Additional readings*

- Simon (2011), [Limited Diagnosticity of Criminal Trials](https://www.marcellodibello.com/innocent/resources/Simon-Diagnosticity2011.pdf)


- Sangero and Halpert (2007), [Why a Conviction Should Not Be Based on a Single Piece of Evidence](https://www.marcellodibello.com/innocent/resources/sangero.pdf)

## 11/21

**Week 14:  Workshop**


Students will workshop their research papers by giving short presentations accompanied by either a handout or slides and followed by Q&A. Each presenter will have 30 minutes to be split between the presentation itself and Q&A at the discretion of the presenter. The list of presenters is TBA.  


## 11/28

**Week 15: Thanksgiving break**


## 12/5

**Week 16: Conclusion**

What have we learned? What questions are still left open and 
what lines of inquiry seemed more promising? Is legal probabilism 
a good idea or a big mistake?

*Primary readings*

- Tribe, [Trial by Methmatics](https://www.maths.ed.ac.uk/~v1ranick/dreyfus/tribe.pdf) 

- Hedden and Colyvan (2019), [Legal Probabilism: A Qualified Defence](http://www.colyvan.com/papers/legalprobabilism) 

*Additional readings*

- Harcourt (2018), [The Systems Fallacy: A Genealogy and Critique of Public Policy and Cost-Benefit Analysis](https://scholarship.law.columbia.edu/faculty_scholarship/2632/)


**Third essay or research paper due December 5th**

# Title IX 

Title IX is a federal law that provides that no person be excluded on the basis of sex from participation in, be denied benefits of, or be subjected to discrimination under any education program or activity.  Both Title IX and university policy make clear that sexual violence and harassment based on sex is prohibited.  An individual who believes they have been subjected to sexual violence or harassed on the basis of sex can seek support, including counseling and academic support, from the university.  If you or someone you know has been harassed on the basis of sex or sexually assaulted, you can find information and resources at https://sexualviolenceprevention.asu.edu/faqs. 

 

As a mandated reporter, I am obligated to report any information I become aware of regarding alleged acts of sexual discrimination, including sexual violence and dating violence.  ASU Counseling Services, https://eoss.asu.edu/counseling is available if you wish to discuss any concerns confidentially and privately. ASU online students may access 360 Life Services, https://goto.asuonline.asu.edu/success/online-resources.html.